Semantic integration of physiology phenotypes with an application to the Cellular Phenotype Ontology

### 2012-03-29 trunk/paper/drinking-bioinformatics.tex

\documentclass{bioinfo}
\copyrightyear{2005}
\pubyear{2005}
\usepackage{url}
\usepackage[normalem]{ulem}
\usepackage{multirow}
\usepackage{pifont}
%\newcommand{\tick}{\ding{52}}
\newcommand{\tick}{\checkmark}
\newcommand{\tickNo}{\hspace{1pt}\ding{55}}

@@

PATO allows for the description of affected entities by combining
various ontologies that describe the entities that have been affected,
such as the various anatomical ontologies, GO \cite {Ashburner2000b},
the Cell Type Ontology \cite {Bard2005}, SO \cite {Eilbeck2005} etc
with the various qualities it provides for defining how these entities
were affected.  PATO can be used for annotation either directly in a
so called post-composed (post-coordinated) manner or for providing
formal (logical) definitions (equivalence axioms) to ontologies
containing a set of precomposed (pre-coordinated) phenotype terms. For
instance, to describe the decrease in the length of the sexual cycle
of female animals, we can combine the PATO term \textit{decreased
  duration} ({\tt PATO:0000499}) with the Gene Ontology term
\textit{estrous cycle} ({\tt GO:0042698}), whilst if such a term
existed in a pre-composed ontology (for example the {\tt MP:0009007}
term from the Mammalian Phenotype) it could be used to provide an
equivalence statement between that class and the above PATO-based
description.
\end{methods}

@@

We implement the first two types of abnormality in the CPO. First, as
a subclass of each {\em Abnormality of P} class, we create {\em
  Abnormal duration of P}, which in turn has {\em Increased duration
  of P} and {\em Decreased duration of P} as subclasses. Second, if we
are able to identify {\em inputs} $I(P)$ or {\em outputs} $O(P)$ of
the process $P$ in the formal definitions of the GO, we automatically
generate {\em Abnormality of $I(P)$ in $P$} as well as {\em
  Abnormality of $O(P)$ in $P$}.  The left side of Figure
\ref{fig:overview} illustrates the schema of classes we generate for
single process abnormalities.
\begin{figure*}
  \centering
%  \includegraphics[width=1.2\textwidth, height=.4\textheight, angle=270]{overview.pdf}
  \includegraphics[width=1.1\textwidth]{overview.pdf}
  \caption{Overview over the taxonomic structure of CPO. The structure
    is based on a cellular component class $X$ and the cellular
    processes $P(X)$ in which $X$ is involved.\label{fig:overview}}
\end{figure*}

@@

\section{Acknowledgement}
Funding for RH is provided by the European Commission's 7th Framework
Programme, RICORDO project, grant number 248502. MAH is supported by
Wellcome Trust, grant WT090548MA. HH is supported by the Institute for
Medical Informatics, Statistics and Epidemiology, University of
Leipzig.  Funding for GR is provided by the European Union's Seventh
Framework Programme (FP7/2007-2013) under grant agreement number
258068, EU-FP7-Systems Microscopy NoE.

\bibliographystyle{plain}
\bibliography{/home/leechuck/Documents/papers/bibtex/lc}

>>>>

\documentclass{bioinfo}
\copyrightyear{2005}
\pubyear{2005}
\usepackage{url}
\usepackage[normalem]{ulem}
\usepackage{multirow}
\renewcommand{\cite}{\citep}

@@

PATO allows for the description of affected entities by combining
various ontologies that describe the entities that have been affected,
such as the various anatomical ontologies, GO \cite {Ashburner2000short},
the Cell Type Ontology \cite {Bard2005}, SO \cite {Eilbeck2005} etc
with the various qualities it provides for defining how these entities
were affected.  PATO can be used for annotation either directly in a
so called post-composed (post-coordinated) manner or for providing
formal (logical) definitions (equivalence axioms) to ontologies
containing a set of precomposed (pre-coordinated) phenotype terms. For
instance, to describe the decrease in the length of the sexual cycle
of female animals, we can combine the PATO term \textit{Decreased
  duration} ({\tt PATO:0000499}) with the Gene Ontology term
\textit{Estrous cycle} ({\tt GO:0042698}), whilst if such a term
existed in a pre-composed ontology (for example the {\tt MP:0009007}
term from the Mammalian Phenotype) it could be used to provide an
equivalence statement between that class and the above PATO-based
description.

\subsection{Axioms for physiology phenotypes}
We implement our theory of physiology phenotypes using OWL, a formal
language based on description logics. Using OWL, we formulate axioms
that can be used by automated reasoners to infer additional
information. Automated reasoning and the axioms we provide are
intended to satisfy three aims. First, we use the axioms to infer
information from background knowledge. In particular, we aim to
automatically generate a class hierarchy of physiology phenotypes when
an ontology of physiological processes, such as the GO, is provided as
background knowledge. Our second aim is to provide interoperability
with phenotype ontologies of other domains, including species-specific
phenotype ontologies that have been formalized using the EQ method
\cite{Mungall2010, Gkoutos2005}. Finally, our third aim is to query
physiology phenotypes based on physiological processes that are
affected or based on the way in which they are affected.

Our three aims rely on the possibility of using automated reasoning
over a resulting ontology of physiology phenotypes.  However, OWL is
an expressive formal language, and automated reasoning in OWL has a
high computational complexity (reasoning in OWL 2.0 is
2NEXPTIME-complete). Consequently, due to the exponential increase in
worst-case time complexity for automated reasoning in OWL, we would
not be guaranteed to achieve our aims once we consider more than very
few phenotype classes. In particular, using an ontology of the size of
GO, with more than 22,000 classes for processes, as a foundation for
constructing an ontology of physiology phenotypes would not allow us
to achieve our aims through automated reasoning.

The OWL EL profile is a subset of OWL that significantly decreases the
expressivity of OWL and the resulting time complexity of automated
reasoning \cite{owlprofiles}. Highly efficient automated reasoners
have been developed for OWL EL which are capable of processing very
large ontologies \cite{Kazakov2011}. In order to achieve our aims and
utilize automated reasoning for large ontologies, we limit ourselves
to the OWL EL profile. As a consequence of the restriction to OWL EL,
we cannot make use of negation ({\tt not}), disjunction ({\tt or}),
universal quantification ({\tt forall}) and several other types of
operations in our axioms \cite{owlprofiles}.

The lack of expressivity in OWL EL requires a formulation of axioms so
that the inferences we desire (i.e., the subclass relations resulting
in the ontology's taxonomy) are maintained without using features of
OWL that go beyond OWL EL expressivity. Consequently, we formulate
{\em abnormality} and {\em absence} similarly to current
formalizations of EQ-based phenotype ontologies, without the use of
negation, disjunction or universal quantification
\cite{Mungall2010}. A detailed description of the axioms we implement
is available as supplementary material (Supplement 1).
\end{methods}

@@

We implement the first two types of abnormality in the CPO. First, as
a subclass of each {\em Abnormality of P} class, we create {\em
  Abnormal duration of P}, which in turn has {\em Increased duration
  of P} and {\em Decreased duration of P} as subclasses. Second, if we
are able to identify {\em inputs} $I(P)$ or {\em outputs} $O(P)$ of
the process $P$ in the formal definitions of the GO, we automatically
generate {\em Abnormality of $I(P)$ in $P$} as well as {\em
  Abnormality of $O(P)$ in $P$}.  The left side of Figure
\ref{fig:overview} illustrates the schema of classes we generate for
single process abnormalities.
\begin{figure*}
  \centering
  \includegraphics[width=1\textwidth]{overview.eps}
%  \includegraphics[width=1\textwidth]{overview.pdf}
  \caption{Overview over the taxonomic structure of CPO. The structure
    is based on a cellular component class $X$ and the cellular
    processes $P(X)$ in which $X$ is involved.\label{fig:overview}}
\end{figure*}

@@

\section{Acknowledgement}
We thank Michel Dumontier for valuable discussions about the formal
representation of the axioms.

Funding for RH is provided by the European Commission's 7th Framework
Programme, RICORDO project, grant number 248502. MAH is supported by
Wellcome Trust, grant WT090548MA. HH is supported by the Institute for
Medical Informatics, Statistics and Epidemiology, University of
Leipzig.  Funding for GR is provided by the European Union's Seventh
Framework Programme (FP7/2007-2013) under grant agreement number
258068, EU-FP7-Systems Microscopy NoE.

\bibliographystyle{plain}

\bibliography{/home/leechuck/Documents/papers/bibtex/lc}
